Wildcard: End user modification of web applications through a data table

### 2020-01-04 paper.tex
finished draft 1; edited the PDF credentials too

@@

\author{Geoffrey Litt}
\authorinfo{is bla bla bla}
\author{Daniel Jackson}
\authorinfo{is bla bla bla}
\affiliation{Massachusetts Institute of Technology}

\keywords{programming journal, paper formatting, submission preparation} % please provide 1--5 keywords

@@

In this paper, we introduce Wildcard\footnote{Wildcard was the internal
  pre-release name for Hypercard, which served as an inspiration for
  this project since it promoted software modification by end users and
  was a precursor to the modern Web.}, a browser extension that aims to
make software more malleable by enabling users to tweak web applications
without programming. Wildcard adds a panel to the bottom of a web page
that shows a structured table view of the main data in the page. The
table maintains a bidirectional connection to the original page---when
the user manipulates the table, the original page gets modified, and
vice versa.

@@

\begin{figure}
\centering
\includegraphics{media/architecture.png}
\caption{The architecture of the Wildcard system}
\end{figure}

@@

In order to maximize extensibility, Wildcard is implemented as a small
core program along with several types of plugins: site adapters,
formulas, and cell renderers/editors. The core contains functionality
for displaying the data table and handling user interactions, and the
table implementation is currently built on the
\href{https://handsontable.com/}{Handsontable} Javascript library to
enable rapid prototyping.

Site adapters\ldots.

Formula functions\ldots{}

Cell viewers\ldots{}

\begin{itemize}
\tightlist
\item
  Extension points

  \begin{itemize}
  \tightlist
  \item
    site adapters

    \begin{itemize}
    \tightlist
    \item
      show a snippet of adapter code
    \item
      discuss how easy it is for programmers to make adapters
    \item
      discuss possible future automation of adapter creation
    \item
      Mention the technique of scraping data from AJAX requests
    \end{itemize}
  \item
    formula functions
  \item
    cell viewer/editors
  \end{itemize}
\item
  future implementation goals

  \begin{itemize}
  \tightlist
  \item
    make it easy for programmers to add adapters + plugins, and
    distribute them to users. (Currently all adapters + plugins are part
    of the main Wildcard codebase)
  \end{itemize}
\end{itemize}

@@

The design of Wildcard is grounded in several principles, informed by
prior work and our own experimentation. We hope you find these
principles helpful not only for understanding our prototype, but also
for designing other systems for end user programming.

\emph{Todo: add diagrams illustrating principles}

@@

The Web is an unusually extensible platform. On many other platforms
(e.g.~smartphone operating systems), software is locked down unless
first-party developers explicitly provide hooks for plugins and
interoperability, but on the Web, all client-side code is available for
browser extensions to modify. Application authors can use practices that
make it easier to modify their apps (e.g.~clean semantic markup), or
more difficult (e.g.~code obfuscation), but the default state is
openness. This gives extensions freedom to modify applications in
creative ways that the original developers did not plan for.

@@

\hypertarget{no-magic}{%
\subsection{No magic}\label{no-magic}}

\begin{itemize}
\tightlist
\item
  an ecosystem of programmers collaboratively building a platform for
  end users
\item
  but still empowering end users to do the creative parts, the final
  glue
\item
  less brittle, less confusing and surprising (maybe Chickenfoot or
  Coscripter had this problem? Find a nugget in those papers to support.
  I remember Helena had something about auto-scraping breaking\ldots{}
  contrast with Gmail.Js stability?)
\item
  still leaves the door open for future automation
\item
  maybe ``first party optional'' can be included here?
\end{itemize}

@@

\emph{Note: a lot of this was already covered above; how to deal with
that?}

\begin{itemize}
\tightlist
\item
  Malleable software: Kay, Webstrates
\item
  Instrumental interaction, polymorphic UI
\item
  Web automation: Chickenfoot, CoScripter
\item
  Wrapper induction: Thresher, Helena
\item
  Personal data ownership: SOLID
\item
  Extension helper libraries, e.g.~Gmail.js.
\item
  Doctorow:
  \href{https://www.eff.org/deeplinks/2019/10/adversarial-interoperability/}{adversarial
  interoperability}
\item
  Marmite
\end{itemize}

\hypertarget{future-work}{%
\section{Future work}\label{future-work}}

\begin{itemize}
\tightlist
\item
  limitations / future ideas

  \begin{itemize}
  \tightlist
  \item
    the spreadsheet language is very primitive, can we make it more
    expressive? what are the exact semantics of fetching data from APIs?
    What about making API requests that do mutation, not just fetching
    data?
  \item
    only works when the user is browsing. Should we explore triggers,
    scheduled scraping?
  \item
    only has spreadsheet-style functional transformations. Should we
    explore imperative workflows? Injecting buttons into pages that do
    things? (Eg, imagine a ``save to google maps'' button that you can
    inject into a page)
  \item
    limited options for how to style injected content, could explore
    styling (eg, maybe you can restyle a table cell and the styling is
    reflected when it's injected into the page?)
  \item
    how much do adapters generalize to many sites? We've built X
    adapters but need to make more
  \item
    page boundaries, eg scraping multi page results
  \item
    how do people save scripts and share them with each other? TBD
  \item
    nested data? lean on SIEUFERD?
  \item
    reduce the number of primitives?
  \end{itemize}
\item
  still in early development; note the beta release plan (tentative:
  target public beta availability at the workshop in March?)
\item
  Could explore automated wrapper induction, building on prior work
\item
  Want to get more real usage of the tool and run usability studies
\item
  Include a link to sign up for future updates
\end{itemize}

>>>>

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}

@@

\author{Geoffrey Litt}
\authorinfo{is a PhD student at MIT, researching programming tools for end users and developers.}
\author{Daniel Jackson}
\authorinfo{is a professor in the Department of Electrical Engineering and Computer Science at MIT, associate director of CSAIL, and a MacVicar Fellow. He leads the Software Design Group.}
\affiliation{Massachusetts Institute of Technology}

\keywords{end-user programming, software customization, web browser extensions} % please provide 1--5 keywords

@@

In this paper, we introduce Wildcard, a browser extension that aims to
make software more malleable by enabling users to tweak web applications
without programming. Wildcard adds a panel to the bottom of a web page
that shows a structured table view of the main data in the page. The
table maintains a bidirectional connection to the original page---when
the user manipulates the table, the original page gets modified, and
vice versa.

@@

@@

In order to maximize extensibility by other programmers, Wildcard is
implemented as a small core program along with several types of plugins:
site adapters, formulas, and cell renderers/editors. The core contains
functionality for displaying the data table and handling user
interactions, and the table implementation is built using the
\href{https://handsontable.com/}{Handsontable} Javascript library.

\begin{figure}
\centering
\includegraphics{media/architecture-clean.png}
\caption{The architecture of the Wildcard system}
\end{figure}

Site adapters specify the bidirectional connection between the web page
and its structured data representation.

For extracting data from the page and getting it into structured form,
Wildcard provides ways to concisely and declaratively express scraping
logic. For example, here is a code snippet for extracting the name of an
Airbnb listing. It specifies the field name and some metadata
properties, and then a function that extracts the name div from a row
div.

\begin{Shaded}
\begin{Highlighting}[]
  \OperatorTok{\{}
\NormalTok{    fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,}
\NormalTok{    readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}
\NormalTok{    type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}
\NormalTok{    el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
  \OperatorTok{\}}
\end{Highlighting}
\end{Shaded}

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter. To support sorting and
filtering, the adapter specifies how to find the divs representing data
rows, and the core platform sorts and filters the DOM to reflect the
table state. User annotations are handled similarly. Finally, site
adapters can also optionally implement site-specific actions (like
favoriting in Airbnb) in normal Javascript code that controls the page.

So far we have implemented X (\emph{fill in latest number}) site
adapters built into Wildcard. We plan to implement more in the near
future to continue to test the limits of the API, and to encourage third
party programmers to build, maintain, and distribute adapters for
popular sites.

Formulas and cell editors are also implemented as separate modules, so
that additional ones could be contributed by third party programmers.

@@

The design of Wildcard is grounded in several principles, informed by
prior work and our own experimentation. We think these principles can
also more generally inform the design of end user programming,
particularly systems enabling users to modify GUI apps.

@@

The Web is an unusually extensible platform. On many other platforms
(e.g.~smartphone operating systems), software is locked down unless
first-party developers explicitly provide hooks for plugins and
interoperability, but on the Web, all client-side code is available for
browser extensions to modify. Application authors can use practices that
make it easier to modify their apps (e.g.~clean semantic markup), or
more difficult (e.g.~code obfuscation), but the default state is
openness. This gives extensions freedom to modify applications in
creative ways that the original developers did not plan for. At the most
extreme, this can even include modifications like ad blocking that are
hostile to the desires of the original developers (a notion Cory
Doctorow calls
\href{https://www.eff.org/deeplinks/2019/10/adversarial-interoperability/}{adversarial
interoperability}), but there are also many other changes which are not
actively hostile, merely falling outside the range of possibilities
originally imagined by the developer.

@@

\hypertarget{collaboration-over-automation}{%
\subsection{Collaboration over
automation}\label{collaboration-over-automation}}

The architecture of Wildcard requires site adapters to abstract the low
level representation of a web page into a higher-level, structured
representation that is easier for end users to work with. Many web
scraping research systems have either automated this process entirely
using heuristics, or created interactive workflows for end users to
specify the translation. In Wildcard, we take a different approach:
programmers manually code site adapters.

This decision is partially a pragmatic one---we haven't had time to
incorporate sophisticated automation yet, and we may choose to do so in
the future. But it is also a principled decision. Automated approaches
tend to work much of the time but not all of the time
\autocite{bolin2005} \autocite{little2010}. In contrast, human
programmers have been able to maintain very stable scraping libraries,
even for complex sites like Gmail \autocite{streak,talwar2019}. We are
willing to trade off lower site coverage in return for more reliable
results on the sites that are covered, because this builds confidence
for users about where they can expect Wildcard to work.

We see this architecture as a form of collaboration between programmers
and end users. In the wild, spreadsheets are built collaboratively by
users with different levels of technical expertise: some users write
simple formulas, while others help write more complex formulas or even
macros \autocite{nardi1990}. Even though some parts of the collaboration
require more expertise, even users with little expertise are still
empowered to perform much of the work themselves, which differs
radically from custom software development where coding expertise is
table stakes for making any contributions. Similarly, we envision
multiple layers of Wildcard usage depending on technical expertise:

\begin{itemize}
\tightlist
\item
  only using sorting/filtering/annotation, with no computation of any
  kind
\item
  writing simple formulas to fetch new data
\item
  composing complex formulas that perform more advanced computation
\item
  using Javascript to create or maintain site adapters, formulas, or
  cell editors
\end{itemize}

@@

Many projects have explored end user customization of web applications,
but it has not become a common activity in the wild. We suspect that so
far no system has found the right combination of utility, ease of use,
and robustness required for widespread adoption. Wildcard aims to build
on lessons learned from this body of work, employing a novel approach of
using a familiar spreadsheet table to go about the task of modifying the
UI of an existing web application. Here we describe some of the areas of
related work, and how Wildcard builds on them.

\hypertarget{malleable-software}{%
\subsection{Malleable software}\label{malleable-software}}

In the broadest sense, Wildcard is inspired by systems aiming to make
software into a dynamic medium where end users frequently create and
modify software to meet their own needs, rather than accepting pre-built
applications built by programmers. These systems include Smalltalk
\autocite{kay1977}, Hypercard \autocite{hypercard2019}, Boxer
\autocite{disessa1986}, Webstrates \autocite{klokmose2015}, and
Dynamicland \autocite{victor}. (The project's name Wildcard comes from
the internal pre-release name for Hypercard, which promoted software
modification by end users and was a precursor to the modern Web.)

These projects generally require building new software from scratch for
that environment. This provides a great deal of flexibility, but also
means they are not compatible with existing software and require
wholesale adoption. In contrast, Wildcard takes a pragmatic approach of
enabling people to tweak the software they already use.

\hypertarget{web-augmentation}{%
\subsection{Web augmentation}\label{web-augmentation}}

Wildcard's goals are closely shared with other browser extensions which
provide panels to augment and modify a website in the context of using
the site.

Sifter \autocite{huynh2006} enables users to sort and filter lists of
data on web pages by converting the data into a structured format,
providing a result similar to Wildcard's sort and filter functionality.
The mechanism is different, though: Sifter uses a combination of
automated heuristics and interactive user feedback to extract data,
whereas Wildcard currently relies on programmers creating wrappers for
extracting structured data, probably leading to higher quality
extraction but on fewer sites. Also, Wildcard aims for much broader
functionality than just sorting and filtering, and shows the structured
table view of the data to facilitate these other interactions, whereas
Sifter just shows sort controls and never reveals the underlying data
table.

``Sloppy programming'' \autocite{little2010} tools like Chickenfoot
\autocite{bolin2005} and Coscripter \autocite{leshed2008} enable users
to create scripts that perform actions like filling in text boxes and
clicking buttons, without directly interacting with the DOM. Users
express the desired page elements in natural, informal terms
(e.g.~writing ``the username box'' to represent the textbook closest to
the label ``username''), and then using heuristics to determine which
elements most likely match the user's intent. This approach allows for
expressing a wide variety of commands with minimal training, but it also
has downsides. It is difficult to know whether a command will
consistently work over time (in addition to changes to the website,
changes to the heuristics can also cause problems), and it is not easy
for users to discover the space of possible commands.

Wildcard offers a sharp contrast to sloppy programming, instead choosing
to expose a high degree of structure through the familiar spreadsheet
table, which offers the reverse set of tradeoffs: more robust behavior,
but in a narrower domain. Wildcard offers more consistency (e.g.,
clicking a sort header will always work correctly as long as the site
adapter is maintained) and offers clearer affordances for what types of
actions are possible (or, crucially, what actions are \emph{not}
possible, which is very useful for users to know). On the other hand,
Wildcard cannot offer coverage of all websites, and has a narrower range
of possible actions than sloppy tools. Still, we hope that with enough
site adapters and formulas, these downsides can be mitigated, and that
the benefits of structure outweigh these costs.

\hypertarget{spreadsheet-based-app-builders}{%
\subsection{Spreadsheet-based app
builders}\label{spreadsheet-based-app-builders}}

It is a powerful realization to notice that a spreadsheet can serve as
an end-user-friendly backing data store and computation layer for an
interactive web application. Research projects like Quilt
\autocite{benson2014}, Gneiss \autocite{chang2014}, and Marmite
\autocite{wong2007}, as well as commercial tools like Airtable Blocks
\autocite{zotero-79} and Glide \autocite{zotero-81} allow users to view
data in a spreadsheet table, compute over the data using formulas, and
connect the table to a GUI. Because many users are already familiar with
using spreadsheets, this way of creating applications tends to be much
easier than traditional software methods; for example, in a user study
of Quilt, many people were able to create applications in under 10
minutes, even if they expected it would take them many hours.

Wildcard builds on this powerful idea, but applies it to modifying
existing applications, rather than building new applications from
scratch. For many people, we suspect that tweaking existing applications
provides more motivation as a starting point for programming than
creating a new application from scratch.

An important design decision for tools in this space is how far to
deviate from traditional spreadsheets, and in what ways. Quilt and Glide
use existing web spreadsheet software as a backend, providing maximum
familiarity and compatibility with existing spreadsheets, but other
projects branch out in various ways. Gneiss has its own spreadsheet with
additional features useful for building GUIs. Marmite provides a live
data view that resembles a spreadsheet, but programming is actually done
using a separate data flow pane rather than spreadsheet formulas, which
led to some confusion in a user study among users who expected behavior
more similar to spreadsheets \autocite{wong2007}. Airtable offers a
relational database with formula support in a spreadsheet-style direct
manipulation interface, a combination which has found commercial
success. Wildcard's table is most similar to Airtable, aiming to offer
the structure of a relational table, but to keep the notion of pure
reactive formulas from spreadsheets rather than introducing a separate
programming model outside of the table.

Quilt also raised the hypothetical idea of ``web worksheets'' in
spreadsheets: small HTML UIs inside of a spreadsheet table to help users
edit data or formulas in a more domain-specific interface. This is
similar to Wildcard's notion of cell editors.

\hypertarget{web-scraping-data-extraction}{%
\subsection{Web scraping / data
extraction}\label{web-scraping-data-extraction}}

Web scraping tools focus on extracting structured data out of
unstructured web pages. Web scraping is closely related to the
implementation of Wildcard, but has very different goals: web scraping
extracts static data for processing in another environment, whereas
Wildcard supports modification of the original page by maintaining a
bidirectional connection between the extracted data and the page.

Web scraping tools differ in how much structure they attempt to map onto
the data. Some tools like Rousillon \autocite{chasins2018} extract data
in a minimally structured relational format; other tools like Piggy Bank
\autocite{huynh2005} or Thresher \autocite{hogue2005} more ambitiously
map the data to a Semantic Web schema. In Wildcard, we chose to avoid
rich schemas, in order to reduce the work associated with creating a
site adapter.

\hypertarget{limitations-and-future-work}{%
\section{Limitations and future
work}\label{limitations-and-future-work}}

Wildcard is still in early development; there are still many limitations
to resolve and open questions to explore.

The most important question is whether the combination of features
provided by Wildcard can actually be combined in enough useful ways to
make the system worth using in practice. We plan to test this by first
continuing to privately test the system with our own needs, and
eventually by shipping the system publicly, once the API is stable
enough and we have developed enough site adapters and formulas to
support a critical mass of sites and use cases.

Here are a few of the most pressing limitations in the current system:

\begin{itemize}
\tightlist
\item
  Wildcard only extracts data on the page, which means that subsequent
  pages of results in paginated lists are not included in the table.
  (Sifter explains \autocite{huynh2006} techniques that might help get
  around this.)
\item
  There is no mechanism for end users to express imperative workflows
  (e.g.~``click this button, and then\ldots{}''); they can only write
  formulas that return data and then inject the resulting data into the
  page. While this makes the system simpler, it also may exclude many
  valuable use cases.
\item
  Wildcard's data model expresses a single table at a time, without any
  notion of relationships between tables. A UI like SIEUFERD
  \autocite{bakke2016} might help to visualize joins between tables.
\item
  Only allowing site adapters to be manually coded means that the number
  of supported sites is limited. We may explore also building wrappers
  using automated heuristics.
\end{itemize}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

We live in a time of pessimism about the effects of the internet and
software. At times it seems like there is no way out of the mess, but we
think that one promising solution is to enable more people to have
deeper control over the software they use every day. The Web offers an
open foundation that browser extensions have capitalized on, but even
this has only empowered users within the limited confines of the ideas
that programmers have been able to think of.

In this paper, we have presented Wildcard, a general-purpose browser
extension that gives end users more freedom to mold web applications to
meet their needs. To receive future updates on Wildcard and
notifications about a public release, sign up for the email newsletter.
